\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A Guide for International Students

### 2023-10-19 demystify.tex
updates

@@

\section{How is Your Application Evaluated?}

@@

\begin{commentbox}[Hung:]
I personally don’t care about which schools students attended. There are many other ways to evaluate an applicant.
\end{commentbox}

@@

\section{Getting Admitted}

@@

\subsection{Research Assistant (RA)}
RA is provided through a professor through their own funding so you can work on their project.  
You do not need to teach as an RA, so you can focus on your research. Depending on the professor, RA may be available during the summer. \S\ref{sec:ra-cost} gives more details on RA budget.

\textbf{How to get RA?} When a professor recruits you, they will likely give you RA right away (e.g., when you apply).  A common scenario is that you first get admitted with TA, and then after a year or two find an advisor to support you with RA.

\begin{commentbox}[Vu:]
If you got recruited by a prof. who would give you RA right away, it's very likely you will get admitted.  For example, if a prof., even if not in PhD admission committee, wants to work with and funds you, the PhD admission committee will respect that decision and admit you (unless your application has many red flags).
\end{commentbox}

@@

There are many types of universities in the US (as well as other countries). But here I focus the discussion in the context of obtaining a Ph.D degree in CS.

\subsection{Schools that offer Ph.D. in CS}  Most university will have CS programs. 
However, while many universities offer Ph.D. degree in their CS programs, many do not.  These universities might offer just Bachelor degrees (e.g., BS) and no graduate studies (i.e., no MS or Ph.D degrees), or they must just offer MS programs (but no PhD).

Thus, if your goal is Ph.D. in CS, you will have to filter out schools that do not offer such degree.  This might sound obvious but many students (even those in the US) get confused due to the large number of universities. For example, Penn State in University Park has Ph.D. in CS,  but Penn State in Harrisburg only has BS and MS in CS and Penn State in York only has BS in CS.  On the other hand, multiplication locations of the University of Texas, e.g., Austin, Dallas, Arlington, all have PhDs in CS. Also, some schools might offer PhDs in related fields, but not CS.

\subsubsection{R1, R2, ...}

@@

However, in general, rankings can be superficial and you need to do more research to be informed and make better decision. For example, if you get admissions to several places, you should contact profs. that you're interested in at those place and talk to them. They would be more willing to chat to you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students.

@@

\subsubsection{Tenured, Tenure Track, Teaching,  Assistant Profs., Associate Profs}

@@

    In other fields, e.g. Software Engineering, Vietnamese students face challenges in reaching US professors. Do you have any tips for Vietnamese students who want to connect with US professors and work as research assistants?
    \end{commentbox}

\begin{commentbox}[Xiaokuan]
Whether the student's research interest matches that of the advisor is very important;
if there is a mismatch,
either the student or the advisor has to make compromises,
which often leads to disagreements or conflicts. 
IMO, the advisor should be the one who {\it guides}  students to do research while allowing students to pursue their own interest,
instead of {\it dictating} their research.

@@

\subsection{Should you contact a professor? What to do to get a desired reply?}

@@

First, if you want to contact a prof. to \emph{ask about your admission chance}, please \textbf{don't}. We don't know and can't answer because as explained throughout this document, we don't make individual decisions and might not even be assigned to evaluate your application.  It is the same as sending a paper draft to a journal editor and ask them if your paper has a chance.

@@

\xz{Applying for Ph.D. and contacting a potential Ph.D. advisor is a classic {\bf `why me, why you'} problem, 
similar to looking for a job in a company.
On a high level,
you need to show that you have done your {\it homework}
regarding the professor and the university,
and clearly explain:
1) why do you think you are a good fit in professor A's group?
2) why do you want to be advised by professor A, not B? 
3) why do you want to apply for university X, not Y?
If you don't want to spend time to do your homework,
the chance of getting a reply is close to zero.
}

@@

@@

 Show something that makes you \textbf{stand out}, e.g., do you have a degree or background in \emph{biology} or \emph{music} and want to integrate them with CS? are you a female or a minority in CS (research for "URM minority in CS in the US" on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS? All of these are unique and would get noticed from reviewers.

@@

    Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs at universities in other countries, which might require MS first (recall that US PhD programs do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students often take a ``leave of absence'' for 1--2 semesters to do internship at companies and research labs.

@@

\subsection{How much do \emph{you} cost?}\label{sec:ra-cost}
PhD students often ask why their salary is so low compared to ludicrous grants their advisors get. They also wonder why their offer letters sometime mentioned that their benefits worth way more than what they actually receive (i.e., stipend).  This section aims to shed some light to these questions.

@@

\section{Cultural Differences and Tips}\label{sec:cultural}

This section lists various things related to studying the US that I notice international students often get surprised or confused.

@@

\paragraph{Academic Integrity} Plagiarism and cheating (e.g., exams, coding) is a big no-no in the US.  If you're caught cheating, you will face very heavy consequence and likely be expelled from the university (e.g., after the second time at GMU).   This is very different from many international countries where cheating is common and often tolerated.  Faculty is extremely good at detecting cheating (we have been dealing with these situations so many times over so many years), and \emph{will} report cheating cases.  In short, whatever you do, don't cheat---not worth it.

Here is the typically steps: (i) a faculty suspecting a cheating case will report it to the Office of Academic Integrity (OAI) at the university;  (ii)  OAI will take over and investigate the case; and finally (iii) OAI will make the final decision.  It is important to note that after receiving the report from your prof., OAI \emph{completely} takes over (so begging your professor will not help because they simply are no longer involved and can't do much here).

\paragraph{Illegal Software} Using illegal/cracked software is very common in many countries. But do not install or use them on university computer (e.g., computers given by your department or advisor).  It is unlikely that the university will track you down, but it is the \emph{software company} that will.  They have very sophisticated tools to detect illegal software and will sue your university/department.  Imagine your department or advisor being sued for half a million dollars, and it is you who caused it.   If you need software, ask your advisor or the department.

@@

\paragraph{History} This document was conceived during a lunch with Craig Yu at GMU.  We talked on about why we were not able to attract good Vietnamese international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \S\ref{sec:ranking}) and wished there were a way for international students to know about the US PhD programs (as well as for US faculty to understand more about international students and therefore have better chance of recruiting and working with them). I was also  a member of the large VietPhD group on Facebook and saw many questions from students about PhD programs.  However, most active participant are students (and some few faculty) in non-CS fields or not in US. Like ChatGPT, their answers to CS PhD program in the US could be informative and helpful, but unfortunately not always accurate and sometimes leading to more confusion. So I thought it would be useful to have a document that is specific to CS PhD programs in the US from an insider prospective.

>>>>

\tableofcontents

@@

\section{How is Your Application Evaluated?}\label{sec:evalapps}

@@

% \begin{commentbox}[Hung:]
% I personally don’t care about which schools students attended. There are many other ways to evaluate an applicant.
% \end{commentbox}

@@

\section{Getting Admitted}\label{sec:accepted}

@@

\subsubsection{Research Assistant (RA)}
RA is provided through a professor through their own funding so you can work on their project.  
You do not need to teach as an RA, so you can focus on your research. Depending on the professor, RA may be available during the summer. \S\ref{sec:ra-cost} gives more details on RA budget.

\textbf{How to get RA?} When a professor recruits you, they will likely give you RA right away (e.g., when you apply).  A common scenario is that you first get admitted with TA, and then after a year or two find an advisor to support you with RA. See how to contact a prof. in \S\ref{sec:contact} for research opportunities.

\begin{commentbox}[Vu:]
If you got recruited by a prof., who would give you RA right away, it's very likely you will get admitted.  For example, if a prof., even if not in PhD admission committee, wants to work with and funds you, the PhD admission committee will respect that decision and admit you (unless your application has many red flags).
\end{commentbox}

@@

This section aims to explain different types of universities in the US in the context of obtaining a CS Ph.D degree.

\subsection{Schools that offer Ph.D. in CS}

Most universities the US will have CS programs. 
However, while many universities offer Ph.D. degree in their CS programs, many do not.  These universities might offer just Bachelor degrees (e.g., BS) and no graduate studies (i.e., no MS or Ph.D degrees), or they just offer MS programs (but no PhD).

Thus, if your goal is Ph.D. in CS, you will have to filter out schools that do not offer such a degree.  This might sound obvious but many students (even those in the US) get confused due to the large number of universities. For example, Penn State in University Park has Ph.D. in CS,  but Penn State in Harrisburg only has BS and MS in CS, and Penn State in York only has BS in CS.  On the other hand, multiplication locations of the University of Texas, e.g., Austin, Dallas, Arlington, all have PhDs in CS.

While this can be confusing at first, a little research, e.g., searching for PhD in CS from the school website, will help you find out. All schools listed in \S\ref{sec:ranking} have PhD studies in CS, so you can start there.

%\subsubsection{R1, R2, ...}

@@

However, in general, rankings can be superficial and you need to do more research to be informed and make better decision. For example, if you get admissions to several places, you should consider attending Open Houses and contact profs. that you're interested in at those place and talk to them. They would be more willing to chat with you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students. See more in \S\ref{sec:accepted} on what to do after getting admitted.

@@

%\subsubsection{Tenured, Tenure Track, Teaching,  Assistant Profs., Associate Profs}

@@

    In other fields, e.g. Software Engineering, Vietnamese students face challenges in reaching US professors. Do you have any tips for Vietnamese students who want to connect with US professors and work as research assistants? \textbf{Vu}: see sec ~\ref{sec:contact} to contact a professor for research opportunities.
    \end{commentbox}

\begin{commentbox}[Xiaokuan:]
Whether the student's research interest matches that of the advisor is very important;
if there is a mismatch,
either the student or the advisor has to make compromises,
which often leads to disagreements or conflicts. 
IMO, the advisor should be the one who {\it guides}  students to do research while allowing students to pursue their own interest,
instead of {\it dictating} their research.

@@

\subsection{Should you contact a professor? What to do to get a desired reply?}\label{sec:contact}

@@

First, if you want to contact a prof. to \emph{ask about your admission chance}, please \textbf{don't}. We don't know and can't answer because as explained in \S\ref{sec:evalapps}, we don't make individual decisions and might not even be assigned to evaluate your application.  It is the same as sending a paper draft to a journal editor and ask them if your paper has a chance.

@@

@@

\begin{commentbox}[Xiaokuan]
    Applying for Ph.D. and contacting a potential Ph.D. advisor is a classic {\bf `why me, why you'} problem, 
similar to looking for a job in a company.
On a high level,
you need to show that you have done your {\it homework}
regarding the professor and the university,
and clearly explain:
1) why do you think you are a good fit in professor A's group?
2) why do you want to be advised by professor A, not B? 
3) why do you want to apply for university X, not Y?
If you don't want to spend time to do your homework,
the chance of getting a reply is close to zero.
\end{commentbox}

@@

 Show something that makes you \textbf{stand out}, e.g., do you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS? are you a female or a minority in CS (research for "URM minority in CS in the US" on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS? All of these are unique and would get noticed from reviewers.

@@

    Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs in other countries, which might require MS first (recall that US PhD programs do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students often take a ``leave of absence'' for 1--2 semesters to do internship at companies and research labs.

@@

\subsection{How much do \emph{you} cost?}\label{sec:ra-cost}
PhD students often ask why their salary is so low compared to ludicrous grants their advisors get. They also wonder why their offer letters sometime mentioned that their benefits appear better more than what they actually receive (i.e., stipend).  This section aims to shed some light to these questions.

@@

\section{Cultural Differences and Other Issues}\label{sec:cultural}

This section lists various issues that international students should pay attention to.

@@

\paragraph{Academic Integrity} Plagiarism and cheating (e.g., exams, coding) is a BIG no-no in the US.  If you're caught cheating, you will face very heavy consequence and likely be expelled from the university (e.g., after the second time at GMU).   This is quite different from many international countries where cheating is common and often tolerated.  Faculty is extremely good at detecting cheating (we have been dealing with these situations so many times over so many years), and \emph{will} report cheating cases.  In short, whatever you do, don't cheat---not worth it.

Here is the typically steps: (i) a faculty suspecting a cheating case will report it to the Office of Academic Integrity (OAI) at the university;  (ii)  OAI will take over and investigate the case; and finally (iii) OAI will make the final decision.  It is important to note that after receiving the report from your prof., OAI \emph{completely} takes over and makes decision.  This means begging your professor will not help because they simply are no longer involved in the case.

\paragraph{Illegal Software} Using illegal/cracked software is very common in many countries (and even in the US). However, \emph{do not} install or use them on university computers (e.g., those given by your department or advisor).  It is unlikely that the university will track you down, but it is the \emph{software company} that will.  They have very sophisticated tools to detect illegal software and will sue your university/department.  Imagine your department or advisor being sued for half a million dollars, and it is \emph{you} who caused it.  If you need software, ask your advisor or the department.

@@

\paragraph{History} This document was conceived during a lunch with Craig Yu at GMU.  We talked on about why GMU were not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \S\ref{sec:ranking}) and wished there were a way for international students to know about the US PhD programs (as well as for US faculty to understand more about international students and therefore have better chance of recruiting and working with them). I was also  a member of the large VietPhD group on Facebook and saw many questions from students about PhD programs.  However, most active participant are students in non-CS fields or not in US. Like ChatGPT, their answers to CS PhD program in the US could be informative and helpful, but unfortunately not always accurate and sometimes leading to more confusion. So I thought it would be useful to have a document that is specific to CS PhD programs in the US from an insider prospective.
